---
title: "Modelling strategy for estimating apparent turnover under enhanced mineral weathering (EMW)"
author: "Teneille Nel"
format: html
editor: visual
---

## Overview

Teneille's full project repository is available at <https://github.com/Teneille97/rocmics>.

We aim to estimate treatment-induced differences in apparent soil organic matter (SOM) turnover under enhanced mineral weathering (EMW), using:

-   Baseline carbon stocks ($t_0$ = May 2023, rock application)
-   Two measured time points ($t_1$ = Oct 2025, $t_2$ = Feb 2026)
-   Bulk soil + three density fractions
-   C stocks and $\Delta^{14}C$
-   3 replicates per treatment
-   3 years since initial pre-treatment C stocks measurement and latest sampling

By treating density fractions as operational pools, fitting transient 1-pool models, and validating against bulk radiocarbon mass balance, we can estimate treatment-induced differences in apparent turnover time without over-parameterizing a multi-pool system. The 3-year divergence window provides the radiocarbon signal necessary to constrain turnover differences for parameter identifiability, while the Oct 2025 → Feb 2026 sampling interval constrains short-term dynamics.

## Conceptual framework & data preparation

**Computational goal**

Each fraction is treated as an operational single pool; this avoids underdetermined multi-pool parameterization.

-   Change in C stocks: $dC/dt = I - kC$
-   Radiocarbon mass balance (Fraction Modern): $d(CF)/dt = I*F~atm~(t) - k*C*F - \lambda *C*F$

**R packages and functions**

-   `Base R/ dplyr`: Use mutate() to convert measured $\Delta^{14}C$ to Fraction Modern ($F$) and structure data (long format): $F = (\Delta^{14}C/1000) + 1$
-   `SoilR`: Load atmospheric radiocarbon curves (e.g., IntCal20 or Hua2013) using bind.C14curves().
-   `rcarbon::calBP()` or IntCal datasets for $F_{atm}(t)$ interpolation

```{r}
#| eval: false

F14C <- (delta14C / 1000) + 1
```

## Forward simulation (3-year history $t_0~to~t_1$)

**Computational goal**

Solve differential equations from known 2023 baseline to 2025 observation (constraint on parameters)

-   Solution (forward): $C(t_1) = C(t_{0})e^{-kt}+(I/k)(1-e^{-kt})$

-   Radiocarbon (integral): $C(t_1)F(t_1) = C(t_0)F(t_0)e^{-(k+\lambda)T}+\int_{a}^{b} [I*F_{atm}(t)*e^{-(k+\lambda)(t_1-t)}] \mathrm{d}t$

**R packages and functions**

-   `SoilR::OnepModel14()`: Create one-pool model object that tracks $^{14}C$ and handles atmospheric input curve $F_{atm}(t)$
-   `SoilR::getC()` and `getF14()`: Extracts carbon stock and radiocarbon values from model object at specific time steps

**Example code**

```{r}
#| eval: false

mod <- OnepModel(t = seq(2023, 2026, length.out = 100), In = I_vec, k = k_init)
fit <- getFitted(mod)
sim_C <- getC(mod)
```

## Cost function to fit k (and possibly I)

**Computational goal**

Define cost function ($J$) that quantifies difference between model predictions and observations at sampling $t_1$ (Oct) and $t_2$ (Feb). Minimize $J$ to find optimal turnover rate ($k$) and input ($I$) for each fraction.

-   Cost function: $J = w_1(C_{t1}^{obs}-C_{t1}^{mod})^2+w_2(C_{t1}^{obs}-C_{t1}^{mod})^2+w_3(C_{t2}^{obs}-C_{t2}^{mod})^2+w_4(C_{t2}^{obs}-C_{t2}^{mod})^2$

**R packages and functions**

-   `FME::modFit()`: Fits models, wraps optimization algorithms e.g. Nelder-Mead, Levenberg-Marquardt, objective J (sum squared residuals for C and F at time $t_1/t_2$); include weights with 'wt' argument
-   `FME::modCost()`: Calculates goodness of fit (weighted sum of squared residuals) between model output and observed df
-   `FME::sensFun()`: Calculates sensitivity of model output to parameters $k$ and $I$ to check for collinearity or identifiability issues before fitting; if I is fixed, bound k \> 0. Use optim() from base R for simpler cases

**Example code**

```{r}
#| eval: false
#| 

# Objective function
objective <- function(p) {
  k <- p["k"]; I <- p["I"]
  # Run forward sim (from step 3) to get C_mod, F_mod at t1/t2
  #...
  sum(c(w1 * (C_t1_obs - C_t1_mod)^2, w2 * (F_t1_obs - F_t1_mod)^2,
             w3 * (C_t2_obs - C_t2_mod)^2, w4 * (F_t2_obs - F_t2_mod)^2))
}

fit <- modFit(f = objective, p = c(k = 0.1, I = 1), 
                   lower = c(k = 0, I = 0), method = "Marq")
     summary(fit)  # Get fitted k and SE
```

## Derive turnover and statistical analysis

**Computational goal**

Convert $k$ to turnover time ($\tau$) and analyze treatment effects across replicates. Account for replicate variability $\epsilon_{ij}$ and treatment means ($\mu_{treatment}$)

-   Treatment effect: $\tau = 1/k_{ij} = \mu_{treatment} + \epsilon_{ij}$

**R packages and functions**

-   `lme4::lmer()`: Fits linear mixed-effects models to estimate frequentist mixed-effects of treatment on $\tau$ while treating replicates as random effects ($1 | replicate$)
-   `brms::brm()`: Bayesian analysis recommended if sample size is small; allows fitting hierarchical model with priors (e.g., Half-Cauchy for $\sigma$, lognormal for $k$) to get credible intervals for treatment differences
-   `emmeans::emmeans()`: Calculates estimated marginal means and performs pairwise contrasts between treatments based on mixed model objects
-   `propagate::propagate()`: Propagate uncertainty

**Example code**

```{r}
#| eval: false


k_est <- coef(fit)["k"]
tau <- 1 / k_est
#error propagation
propagate(expr = 1/k, data = list(k = coef(fit)))
# Delta method approximation
tau_se <- tau * (se_k / k_est)  

#mixed effects on estimated \tau
# Data frame: tau_df with columns tau, treatment, replicate
mod_hier <- lmer(tau ~ treatment + (1 | replicate), data = tau_df)
summary(mod_hier)

#Bayesian fits with priors
prior_k <- prior(exponential(0.1), class = "b", coef = "treatmentControl")
mod_bayes <- brm(k ~ treatment + (1 | replicate), data = k_df, prior = prior_k)  
posterior_summary(mod_bayes)

```

## Validation (bulk reconstruction)

**Computational goal**

Check if sum of independently fitted fractions match observed bulk soil data for both stocks and radiocarbon abundance

-   Validation: $C_{bulk} = \Sigma C_i$
-   Weighted mean $F_{bulk}$ = $\Sigma (C_i*F_i)/\Sigma C_i$

**R packages and functions**

-   `dplyr::group_by()` and `summarise()` to aggregate fitted fraction stocks into calculated bulk stocks
-   `hydroGOF::rmse()`: Calculate root mean square error between reconstructed bulk and measured bulk to assess model performance

## Can EMW affect transfer between SOM pools?

**Possibility of increased transfer from fast to slow pools**

-   Increased mineral surface area (weathering products and fresh reactive surfaces) for adsorption of DOM and microbial products form MAOM, especially in coarse or mineral-limited control soils
-   Nutrients released by ERW can potentially increase microbial CUE which converts more biomass into necromass, primary precursor of MAOM, especially in nutrient-limited control sols
-   Higher plant productivity increases root-derived C inputs which are more accessible for microbes to process, resulting in greater transfer to MAOM
-   Increased pH can enhance microbial transformation and transfer

**Possibility of reduced transfer from fast to slow pools**

-   If pH is too high it can reduce SOM sorption in variable-charge mineral type soils
-   C is lost rapidly as CO₂ (low CUE systems)
-   Weathering products & microbial activity can enhance soil aggregation and reduce interaction between oPOM and other pools

**Radiocarbon implications**

-   Transfer creates “young-but-stable” carbon: Slow pool receives recent carbon via transfer so it can be larger in size, but younger than expected
-   Transfer coefficient \>0 improves model fit
-   Transfer coefficient of treatments \> control means that ERW increased stabilization

**Models**

Transfer term (fast → slow), k\_{fs} and term for fraction of fast-pool losses transferred to slow pool ($\alpha$).

**Example code**

```{r}
#| eval: false

# Fast pool stocks
#\frac{dC_f}{dt} = I_f - k_f C_f - k_{fs} C_f
dCf = I_f - loss_fast_total
loss_fast_total = k_f * Cf
flux_fast_to_CO2  = (1 - alpha) * loss_fast_total

# Slow pool stocks
#\frac{dC_s}{dt} = I_s + k_{fs} C_f - k_s C_s
dCs = I_s + flux_fast_to_slow - flux_slow_to_CO2
flux_fast_to_slow = alpha * loss_fast_total
flux_slow_to_CO2 = k_s * Cs

# 14C tracking: transfer flux inherits fast pool signature: 
F14C_flux_fast_to_slow = F14Cf
d(F14Cf * Cf) = I_f * F14C_input - loss_fast_total * F14Cf
d(F14Cs * Cs) = I_s * F14C_input +  flux_fast_to_slow * F14Cf - flux_slow_to_CO2 * F14Cs
```

**SoilR two pool system**

$$\frac{dC}{dt} = I + A \cdot C(t)$$

Where $A$ is the compartmental matrix. For fast-to-slow transfer:

-   Diagonal elements ($a_{ii}$): Total output rate from pool $i$.
-   Off-diagonal elements ($a_{ij}$): Transfer rate from pool $j$ to pool $i$.
-   $a_{11} = -k_f$ (Total loss from fast pool)
-   $a_{21} = \alpha \cdot k_f$ (Fraction of fast pool loss moving to slow pool)
-   $a_{22} = -k_s$ (Loss from slow pool)

**Example code**

```{r}
#| eval: false
# 1. Setup Time and Atmospheric 14C
t_start <- 2023
t_end <- 2026
t_seq <- seq(t_start, t_end, by = 1/12) # Monthly steps for precision

# 2. Define Parameters
# p = c(k_f, k_s, alpha, I_f, I_s)
p <- c(k_f = 0.5, k_s = 0.05, alpha = 0.2, I_f = 100, I_s = 10)

# 3. Create the Compartmental Matrix A
# Matrix format: A[row, col] -> flow from col to row
A <- matrix(nrow = 2, ncol = 2, byrow = TRUE, c(
  -p["k_f"],                0,          # Pool 1 (Fast)
   p["alpha"] * p["k_f"],  -p["k_s"]    # Pool 2 (Slow) receives alpha*k_f
))

# 4. Run the Model
model <- GeneralModel_14(
  t = t_seq,
  A = A,
  ivList = c(C_f0, C_s0),  # Initial stocks from May 2023
  initialFilters = c(F14_f0, F14_s0), # Initial 14C
  inputFluxes = c(p["I_f"], p["I_s"]),
  inputFc = Hua2013, # Or your preferred curve
  pass = TRUE
)

# 5. Extract results for Fitting
C_total <- getC(model)
F14_total <- getF14(model)

# Updating the Cost Function (FME)
# Evaluate the fast pool, sow pool, and bulk soil simultaneously.

objective_coupled <- function(p) {
  # Constraints: alpha must be 0-1, ks < kf
  if(p["alpha"] < 0 | p["alpha"] > 1 | p["k_s"] >= p["k_f"]) return(Inf)
  
  # 1. Run coupled model (as above)
  # ... model code ...
  
  # 2. Extract predicted values at specific time points (Oct 25, Feb 26)
  # preds <- ... 
  
  # 3. Calculate Multi-target Cost
  cost <- modCost(model = preds, obs = obs_fast, err = "sd")
  cost <- modCost(model = preds, obs = obs_slow, cost = cost, err = "sd")
  return(cost)
}

fit_coupled <- modFit(f = objective_coupled, p = start_params, method = "Marq")
```

**Check if 14C data can distinguish between transfer rate and turnover rates**

-   Run the optimization with $\alpha$ fixed at 0.1 or 0.2 (common literature values) to stabilize $k_f$ and $k_s$.
-   Use FME::sensFun to see if your 14C signal is strong enough. If the sensitivity of $\alpha$ is low, data cannot support a coupled model and should revert to independent pool approach.
-   Bayesian: In brms, set a tight prior on $\alpha$ based on the expected mineral stabilization effect of EMW treatment

## Ideas for incorporation of CUE

**How can carbon use efficiency data be incorporated**

-   CUE = C (microbial growth) / C (uptake)
-   Respiration = 1 - CUE
-   Carbon use efficiency increases biomass production, which is a precursor to stabilized SOM
-   However not all microbial biomass is stabilized (may be recycled, respired etc) - Therefore, we could express the fraction of microbial products that is stabilized into MAOM, with a coefficient $f_{stabilization}$
-   Independently constraining microbial efficiency and inferred mineral stabilization efficiency improves both identifiability and mechanistic insight

**Updating the model with CUE**

-   Replace the transfer fraction with a term that is a product of CUE and stabilization efficiency

*Option A (if build initial models without CUE and incorporate it later):*

-   Instead of fitting alpha, decompose $\alpha$ into:

$\alpha$ = CUE x $f_{stabilization}$

where: CUE = measured $f_{stabilization}$ = fitted stabilization efficiency

```{r}
#| eval: false
alpha = CUE * $f_{stab}$
```

*Option B (if we wait until we have CUE data):*

-   Fast-pool loss becomes:

```{r}
#| eval: false
C_loss = k * C
C_to_growth = CUE * C_loss
C_to_resp = (1 - CUE) * C_loss
C_to_slow = $f_{stab}$ * C_to_growth
```

-   Since CUE is measured in short-term incubation, it is safer to be treated as a process constraint in the time-integrated model, not as a direct flux
-   Therefore, Option A probably better (only controls capacity for transfer, letting $f_{stat}$ include temporal variability) than Option B (flux partitioning of short-term CUE is applied uniformly over long-term)

**How data could be interpreted**

-   If CUE ↑ but no increase in MAOM formation: stabilization limited
-   If MAOM ↑ but CUE unchanged: mineral control dominates
-   If both ↑: strong coupling

**Considerations**

-   CUE is measured from bulk soil, but transfer mostly feeds MAOM, so may need to make assumptions about where growth occurs
-   Assuming realistic ranges of CUE (0.2-0.6) and alpha (0.02-0.3), $f_{stab}$ can vary an order of magnitude (0.05-0.5), therefore alpha is much more sensitive to $f_{stab}$ than to CUE
-   Incubation conditions: we are modelling dynamics of decomposed SOM, not added substrate, so avoid strong priming conditions (use field moisture and temperature, and not high glucose additions) unless signal is not detectable, and avoid longer incubation (community shifts)

**Measure respiration and microbial biomass for validation and interpretation**

-   Scenario 1: CUE ↑, respiration ↓, biomass ↑, more C retained in microbial products → higher transfer potential

-   Scenario 2: CUE ↑, respiration ↓, biomass unchanged, growth occurs but turnover is fast → limited stabilization

-   Scenario 3: CUE unchanged, biomass ↑, MAOM ↑, stabilization efficiency ↑ → f stab effect

**Three-pool model with CUE**

fPOM → oPOM → MAOM

↘────────────→

$\alpha$ -\> {$a_{fo}$, $a_{fm}$, $a_{om}$} 

and 

$a_{fo}$ + $a_{fm}$ = CUE 

where fo, fm and om represent transfer from fPOM to oPOM, fPOM to MAOM and oPOM to MAOM

For each pool i, assimilated = $CUE * k_iC_i$ 
and respiration = $(1-CUE)k_iC_i$

Introduce partitioning coefficient $p$ (fraction of fPOM -> MAOM): 

$p_{fo}$ + $p_{fm}$ = 1

$transfer_{i->j} = CUE * p_{ij} * k_i C_i$

$a_{fm}$ = CUE * $p$

$a_{fo}$ = CUE * $(1-p)$

$a_{om}$ = CUE * $p_{om}$ 

where $p_{om}$ is estimated (fitted?)

Link back $\alpha$ = CUE * $f_stab$
with
$f_stab$ = p + (1-p) * $p_{om}$ 

**SoilR matrix**

$$C = \begin{bmatrix}
   C_f \\
   C_o \\
   C_m
   \end{bmatrix}$$

$$A = \begin{bmatrix}
   -k_f & 0 & 0 \\
   CUE(1-p)k_f & -k_o & 0 \\
   CUEpk_f & CUEp_{om}k_o & -k_m
   \end{bmatrix}$$

Row 1: $-k_fC_f$ = total fPOM loss via decomposition
Row 2: $CUE(1-p)k_fC_f$ = fraction of fPOM routed to oPOM
Row 3: $CUE*pk_fC_f$ = fraction of fPOM stabilized as MAOM, $CUE*p_om*k_oC_o$ = fraction of oPOM stabilized as MAOM

Note MAOM has very small or negligible $k_m$, could set as zero