---
title: "Modelling strategy for estimating apparent turnover under enhanced mineral weathering (EMW)"
author: "Teneille Nel"
format: html
editor: visual
---

## Overview

Teneille's full project repository is available at <https://github.com/Teneille97/rocmics>.

We aim to estimate treatment-induced differences in apparent soil organic matter (SOM) turnover under enhanced mineral weathering (EMW), using:

-   Baseline carbon stocks ($t_0 \approx 2023$, rock application)
-   Two measured time points ($t_1$ = Oct 2025, $t_2$ = Feb 2026)
-   Bulk soil + three density fractions
-   C stocks and $\Delta^{14}C$
-   3 replicates per treatment
-   3 years since initial pre-treatment C stocks measurement and latest sampling

By treating density fractions as operational pools, fitting transient 1-pool models, and validating against bulk radiocarbon mass balance, we can estimate treatment-induced differences in apparent turnover time without over-parameterizing a multi-pool system. The 4-year divergence window provides the radiocarbon signal necessary to constrain turnover differences for parameter identifiability, while the Oct 2025 → Feb 2026 sampling interval constrains short-term dynamics.

## Conceptual framework & data preparation

**Computational goal**

Each fraction is treated as an operational single pool; this avoids underdetermined multi-pool parameterization.

-   Change in C stocks: $dC/dt = I - kC$
-   Radiocarbon mass balance (Fraction Modern): $d(CF)/dt = I*F~atm~(t) - k*C*F - \lambda *C*F$

**R packages and functions**

-   `Base R/ dplyr`: Use mutate() to convert measured $\Delta^{14}C$ to Fraction Modern ($F$) and structure data (long format): $F = (\Delta^{14}C/1000) + 1$
-   `SoilR`: Load atmospheric radiocarbon curves (e.g., IntCal20 or Hua2013) using bind.C14curves().
-   `rcarbon::calBP()` or IntCal datasets for $F_{atm}(t)$ interpolation

**Example code**

```{r}
#| eval: false

F14C <- (delta14C / 1000) + 1
```

## Forward simulation (3-year history $t_0~to~t_1$)

**Computational goal**

Solve differential equations from known 2023 baseline to 2025 observation (constraint on parameters) - Solution (forward): $C(t) = C(t\_{0}*e^{-kt}+(I/k)*(1-e^{-kt})$

-    Radiocarbon (integral): $C(t_1)F(t_1) = C(t_0)F(t_0)e^{-(k+\lambda)T}+\int_{a}^{b} [I*F_{atm}(t)*e^{-(k+\lambda)(t_1-t)}] \mathrm{d}t$

**R packages and functions**

-   `SoilR::OnepModel14()`: Create one-pool model object that tracks $^{14}C$ and handles atmospheric input curve $F_{atm}(t)$
-   `SoilR::getC()` and `getF14()`: Extracts carbon stock and radiocarbon values from model object at specific time steps

**Example code**

```{r}
#| eval: false


mod <- OnepModel(t = seq(2023, 2026, length.out = 100), In = I_vec, k = k_init)
fit <- getFitted(mod)
sim_C <- getC(mod)
```

## Cost function to fit k (and possibly I)

**Computational goal**

Define cost function ($J$) that quantifies difference between model predictions and observations at sampling $t_1$ (Oct) and $t_2$ (Feb). Minimize $J$ to find optimal turnover rate ($k$) and input ($I$) for each fraction.

-   Cost function: $J = w_1(C_{t1}^{obs}-C_{t1}^{mod})^2+w_2(C_{t1}^{obs}-C_{t1}^{mod})^2+w_3(C_{t2}^{obs}-C_{t2}^{mod})^2+w_4(C_{t2}^{obs}-C_{t2}^{mod})^2$

**R packages and functions**

-   `FME::modFit()`: Fits models, wraps optimization algorithms e.g. Nelder-Mead, Levenberg-Marquardt, objective J (sum squared residuals for C and F at time $t_1/t_2$); include weights with 'wt' argument
-   `FME::modCost()`: Calculates goodness of fit (weighted sum of squared residuals) between model output and observed df
-   `FME::sensFun()`: Calculates sensitivity of model output to parameters $k$ and $I$ to check for collinearity or identifiability issues before fitting; if I is fixed, bound k \> 0. Use optim() from base R for simpler cases

**Example code**

```{r}
#| eval: false


# Objective function
objective <- function(p) {
  k <- p["k"]; I <- p["I"]
  # Run forward sim (from step 3) to get C_mod, F_mod at t1/t2
  #...
  sum(c(w1 * (C_t1_obs - C_t1_mod)^2, w2 * (F_t1_obs - F_t1_mod)^2,
             w3 * (C_t2_obs - C_t2_mod)^2, w4 * (F_t2_obs - F_t2_mod)^2))
}

fit <- modFit(f = objective, p = c(k = 0.1, I = 1), 
                   lower = c(k = 0, I = 0), method = "Marq")
     summary(fit)  # Get fitted k and SE
```

## Derive turnover and statistical analysis

**Computational goal**

Convert $k$ to turnover time ($\tau$) and analyze treatment effects across replicates. Account for replicate variability $\epsilon_{ij}$ and treatment means ($\mu_{treatment}$)

-   Treatment effect: $\tau = 1/k_{ij} = \mu_{treatment} + \epsilon_{ij}$

**R packages and functions**

-   `lme4::lmer()`: Fits linear mixed-effects models to estimate frequentist mixed-effects of treatment on $\tau$ while treating replicates as random effects ($1 | replicate$)
-   `brms::brm()`: Bayesian analysis recommended if sample size is small; allows fitting hierarchical model with priors (e.g., Half-Cauchy for $\sigma$, lognormal for $k$) to get credible intervals for treatment differences
-   `emmeans::emmeans()`: Calculates estimated marginal means and performs pairwise contrasts between treatments based on mixed model objects
-   `propagate::propagate()`: Propagate uncertainty

**Example code**

```{r}
#| eval: false


k_est <- coef(fit)["k"]
tau <- 1 / k_est
#error propagation
propagate(expr = 1/k, data = list(k = coef(fit)))
# Delta method approximation
tau_se <- tau * (se_k / k_est)  

#mixed effects on estimated \tau
# Data frame: tau_df with columns tau, treatment, replicate
mod_hier <- lmer(tau ~ treatment + (1 | replicate), data = tau_df)
summary(mod_hier)

#Bayesian fits with priors
prior_k <- prior(exponential(0.1), class = "b", coef = "treatmentControl")
mod_bayes <- brm(k ~ treatment + (1 | replicate), data = k_df, prior = prior_k)  
posterior_summary(mod_bayes)

```

## Validation (bulk reconstruction)

**Computational goal**

Check if sum of independently fitted fractions match observed bulk soil data for both stocks and radiocarbon abundance

-   Validation: $C_{bulk} = \Sigma C_i$
-   Weighted mean $F_{bulk}$ = $\Sigma (C_i*F_i)/\Sigma C_i$

**R packages and functions**

-   `dplyr::group_by()` and `summarise()` to aggregate fitted fraction stocks into calculated bulk stocks
-   `hydroGOF::rmse()`: Calculate root mean square error between reconstructed bulk and measured bulk to assess model performance
