---
title: "atr_aug2026"
format: html
editor: visual
---

## ATR experiment

ATR-FTIR measurements of soil samples were taken before and after extractions of base cations (1M NH4AOc, pH = 7) to evaluate changes in organo-mineral environments between the control and rock powder treatments.

ATR measurements were imported to R and the data was cleaned by trimming to 800-3000 cm $^-1$, baseline correction by asymmetric least squares algorithm (improves signal dampened by noise and spectral drift) and vector normalization (relative peak intensities to account for different sample thickness).

We wanted to investigate whether the following mechanisms of SOM stabilization may be occuring: 

```{r}
#| echo: false
#| message: false
#| label: Table
#| warning: FALSE

library(gt)
library(tibble)

tribble(
  ~`Functional Group`, ~`Chemical Form`, ~`Cation Bridging Mechanism & Relevance`, ~`Organomineral Complexation Mechanisms`,
  "Carboxylate", "-COO^-", "Primary driver (High): Negative electrostatic charge coordinates polyvalent cations.", "Direct ligand exchange (inner-sphere) with surface Fe/Al-OH groups.",
  "Aromatic", "Ar-C=C / Phenolic", "Moderate: Delocalized pi-system coordinates cations via cation-pi interactions.", "Pi-pi stacking, siloxane face adsorption, and ligand exchange if phenolic.",
  "Carbonyl", "-C=O", "Low to Moderate: Dipolar oxygen lone pairs act as weaker Lewis bases.", "Hydrogen bonding with surface hydroxyls (Fe/Al/Si-OH) and dipole interactions.",
  "Aliphatic", "-CH2-, -CH3", "None (Incapable): Lacks charges, lone pairs, or pi-electrons.", "Hydrophobic interactions and van der Waals forces on uncharged surfaces."
) %>%
  gt() %>%
  fmt_markdown(columns = everything())
```

```{r}
#| echo: false
#| message: false
#| label: Spectra before and after
#| warning: FALSE

library(stringr)
library(reshape)
library(data.table)
library(nc)
library(baseline)
library(ggplot2)
library(tidyr)
library(pracma)
library(dplyr)
library(prospectr)
library(broom)
library(here)
library(FSA)
library(car)
library(multcompView)

# ---------------------------------------------------------
# Metadata
# ---------------------------------------------------------
metadata <- read.csv(here("csv_files", "treatment_names.csv"))

metadata <- metadata %>%
  mutate(
    sample = paste0("soil", sample),
    
    # Extract Hühnerberg application rate from Treatment
    Dose = case_when(
      str_detect(Treatment, "^Huhnerberg_") ~ 
        as.numeric(str_extract(Treatment, "(?<=_)\\d+")),
      TRUE ~ NA_real_
    )
  )

metadata <- metadata %>%
  mutate(
    analysis_group = case_when(
      Treatment == "Control" ~ "Control",
      Treatment == "Eifelgold_50" ~ "Eifelgold_50",
      Treatment == "Bolsdorfer_50" ~ "Bolsdorfer_50",
      Treatment == "Lime_2" ~ "Lime_2",
      Treatment == "Huhnerberg_50" ~ "Huhnerberg_50",
      TRUE ~ NA_character_
    )
  )

metadata_analysis <- metadata %>%
  filter(!is.na(analysis_group))

metadata_join <- metadata %>%
  select(
    sample,
    Treatment,
    Basalt_type_exp,
    Dose_resp_exp,
    analysis_group,
    Dose
  )
# ---------------------------------------------------------
# PROCESS SPECTRA
# ---------------------------------------------------------

process_spectra <- function(path){
  
  # ---- import ----
  long_table <- data.table(
    filename = Sys.glob(path)
  )[by = filename,
    j = fread(filename, col.names = c("wavenumber", "absorbance"))
  ]
  
  # ---- metadata ----
  meta <- nc::capture_first_vec(
    long_table$filename,
    sample_no = "soil\\d+",
    rep_no = "[abc]"
  )
  
  meta_long <- cbind(long_table, meta)
  
  meta_long <- meta_long %>%
    mutate(
      wavenumber = round(wavenumber, 0),
      sample = gsub("\\.", "", sample_no),
      sample_rep = paste(sample, rep_no, sep = "_")
    ) %>%
    rename(
      wave = wavenumber,
      abs = absorbance,
      filename = filename
    )
  
  # ---------------------------------------------------------
  # BUILD WIDE MATRIX (raw)
  # ---------------------------------------------------------
  spec_wide <- dcast(meta_long, sample_rep ~ wave, value.var = "abs")
  
  ids <- spec_wide$sample_rep
  spec_mat <- as.matrix(spec_wide[, -1, with = FALSE])
  waves <- as.numeric(colnames(spec_wide)[-1])
  
  # ---------------------------------------------------------
  # BASELINE CORRECTION
  # ---------------------------------------------------------
  bl <- baseline::baseline(spec_mat, method = "als")
  spec_corr <- getCorrected(bl)
  
  corr_mat <- as.data.frame(spec_corr)
  corr_mat$sample_rep <- ids
  
  corr_long <- data.table::melt(
    setDT(corr_mat),
    id.vars = "sample_rep",
    variable.name = "wave",
    value.name = "abs_corr"
  )
  
  corr_long[, wave := as.numeric(as.character(wave))]
  
  # ---------------------------------------------------------
  # DEFINE COMMON GRID HERE
  # ---------------------------------------------------------
  common_grid <- sort(unique(corr_long$wave))
  
  # force consistent grid ordering
  corr_long <- corr_long %>%
    filter(wave %in% common_grid)
  
  
  list(
    raw = meta_long,
    corrected = corr_long,
    peaks = peaks,
    grid = common_grid
  )
}

# ---------------------------------------------------------
# RUN
# ---------------------------------------------------------

before <- process_spectra("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/rocmics/dpt_files/aug2026_dpt_before/*.dpt")
after  <- process_spectra("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/rocmics/dpt_files/aug2026_dpt_after/*.dpt")

# ---------------------------------------------------------
# FILTER REGION
# ---------------------------------------------------------

before_filt <- before$corrected %>%
  filter(wave >= 800, wave <= 3000)

after_filt <- after$corrected %>%
  filter(wave >= 800, wave <= 3000)

# ---------------------------------------------------------
# NORMALIZATION
# ---------------------------------------------------------

normalize_spectra <- function(df){
  
  df %>%
    group_by(sample_rep) %>%
    mutate(
      abs_norm = abs_corr / sqrt(sum(abs_corr^2, na.rm = TRUE))
    ) %>%
    ungroup()
}

before_filt <- normalize_spectra(before_filt)
after_filt  <- normalize_spectra(after_filt)

# ---------------------------------------------------------
# KEEP INDIVIDUAL SCANS FOR INTEGRATION
# ---------------------------------------------------------

before_individual <- before_filt %>%
  mutate(sample = sub("_[a-z]+$", "", sample_rep))

after_individual <- after_filt %>%
  mutate(sample = sub("_[a-z]+$", "", sample_rep))

# ---------------------------------------------------------
# SECOND DERIVATIVE
# ---------------------------------------------------------

compute_second_derivative <- function(df){
  
  scans <- unique(df$sample_rep)
  grid <- sort(unique(df$wave))
  n <- length(grid)
  
  out <- list()
  
  for(scan in scans){
    
    tmp <- df %>%
      filter(sample_rep == scan) %>%
      arrange(wave)
    
    y <- tmp$abs_corr
    
    y2 <- savitzkyGolay(
      matrix(y, nrow = 1),
      p = 3,
      w = 11,
      m = 2
    )[1, ]
    
    
    if(length(y2) != n){
      
      pad <- rep(NA, n)
      
      offset <- floor((n - length(y2)) / 2)
      
      pad[(offset + 1):(offset + length(y2))] <- y2
      
      y2 <- pad
    }
    
    
    out[[scan]] <- data.frame(
      sample_rep = scan,
      sample = unique(tmp$sample),
      wave = grid,
      deriv2 = as.numeric(y2)
    )
  }
  
  bind_rows(out)
}


# plot overlaid sample spectra

samples <- unique(before_individual$sample)
overlay_all <- bind_rows(
  before_individual %>% mutate(dataset = "Before"),
  after_individual %>% mutate(dataset = "After")
) %>%
  mutate(
    sample = factor(
      sample,
      levels = paste0("soil", 1:21)
    )
  )

#### overlay by treatment group 
# Join treatment metadata to individual corrected spectra
overlay_treatment <- overlay_all %>%
  left_join(metadata_join, by = "sample") %>%
  filter(!is.na(analysis_group))

# Average spectrum per treatment group

overlay_treatment_mean <- overlay_treatment %>%
  mutate(dataset = factor(dataset, levels = c("Before", "After"))) %>%
  group_by(dataset, analysis_group, wave) %>%
  summarise(
    mean_abs = mean(abs_corr, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  overlay_treatment_mean,
  aes(x = wave, y = mean_abs, color = analysis_group)
) +
  geom_line(linewidth = 0.8) +
  scale_x_reverse() +
  facet_wrap(~ dataset, ncol = 2) +
  theme_bw() +
  labs(
    x = expression("Wavenumber (cm"^-1*")"),
    y = "Mean Baseline-Corrected Absorbance",
    color = "Treatment Group",
    title = "Average Spectra by Treatment Group"
  ) +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  )

```

Integrated areas were calculated over the spectral ranges associated with functional groups typically found in aliphatics (2898-2976 and 2939-2870 cm$^{-1}$), aromatics (1500-1550 cm$^{-1}$), carboxylates (1380-1425 and 1580-1650  cm$^{-1}$) and carbonyl (1570-1710  cm$^{-1}$) groups. Areas of replicate scans were averaged to yield mean change in areas before and after extraction.


```{r}
#| echo: false
#| message: false
#| label: Integrated peaks
#| warning: FALSE

# ---------------------------------------------------------
# Function to integrate spectral regions per scan
# ---------------------------------------------------------

integrate_bands <- function(df){
  
  bands <- list(
    COO_sym  = c(1380,1425),
    COO_asym = c(1580,1650),
    Silicate = c(1000,1120),
    poorly_crystalline = c(950,1050),
    Carbonate_bend = c(870,880),
    carbonate_stretch = c(1415,1480),
    CH_assym = c(2898,2976),
    CH_sym = c(2839,2870),
    aromatic = c(1500,1550),
    carbonyl = c(1570,1710),
    polysaccharide = c(1030,1080)
  )
  
  
  results <- lapply(unique(df$sample_rep), function(scan){
    
    sub <- df %>%
      filter(sample_rep == scan)
    
    out <- data.frame(
      sample_rep = scan,
      sample = unique(sub$sample)
    )
    
    for(b in names(bands)){
      
      rng <- bands[[b]]
      
      tmp <- sub %>%
        filter(wave >= rng[1],
               wave <= rng[2]) %>%
        arrange(wave)
      
      # avoid errors if region missing
      if(nrow(tmp) > 1){
        area <- trapz(tmp$wave, tmp$abs_norm)
      } else {
        area <- NA
      }
      
      out[[b]] <- area
    }
    
    out
  })
  
  bind_rows(results)
}

# ---------------------------------------------------------
# Calculate integrated areas per rep
# ---------------------------------------------------------
before_scan_areas <- integrate_bands(before_individual)

after_scan_areas <- integrate_bands(after_individual)

# ---------------------------------------------------------
# Calculate combined functional groups per scan
# ---------------------------------------------------------

before_scan_areas <- before_scan_areas %>%
  mutate(
    Aliphatic_total = CH_assym + CH_sym,
    COO_total = COO_sym + COO_asym
  )

after_scan_areas <- after_scan_areas %>%
  mutate(
    Aliphatic_total = CH_assym + CH_sym,
    COO_total = COO_sym + COO_asym
  )

before_scan_areas <- before_scan_areas %>%
  mutate(
    total_area = Silicate + poorly_crystalline +
      carbonate_stretch +
      carbonyl +
      Aliphatic_total
  )

after_scan_areas <- after_scan_areas %>%
  mutate(
    total_area = Silicate + poorly_crystalline +
      carbonate_stretch +
      carbonyl +
      Aliphatic_total
  )
# ---------------------------------------------------------
# Average integrated peak areas across replicate scans
# ---------------------------------------------------------

before_areas <- before_scan_areas %>%
  group_by(sample) %>%
  summarise(
    across(
      where(is.numeric),
      mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  mutate(
    Aliphatic_total = CH_assym + CH_sym,
    COO_total = COO_sym + COO_asym,
    dataset = "before"
  ) %>%
  select(
    -CH_assym,
    -CH_sym,
    -COO_sym,
    -COO_asym
  )


after_areas <- after_scan_areas %>%
  group_by(sample) %>% 
  summarise(
    across(
      where(is.numeric),
      mean,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  mutate(
    Aliphatic_total = CH_assym + CH_sym,
    COO_total = COO_sym + COO_asym,
    dataset = "after"
  ) %>%
  select(
    -CH_assym,
    -CH_sym,
    -COO_sym,
    -COO_asym
  )

before_areas <- before_areas %>%
  left_join(metadata_join, by = "sample")

after_areas <- after_areas %>%
  left_join(metadata_join, by = "sample")

# ---------------------------------------------------------
# Combine and calculate extraction effect
# ---------------------------------------------------------

band_compare <- before_areas %>%
  
  # Rename spectral variables from BEFORE
  rename_with(
    ~ paste0(.x, "_before"),
    -c(sample, Treatment, analysis_group, Dose)
  ) %>%
  
  # Join AFTER spectral data
  left_join(
    after_areas %>%
      select(
        sample,
        Treatment,
        analysis_group,
        everything()
      ) %>%
      select(-Dose) %>%
      rename_with(
        ~ paste0(.x, "_after"),
        -c(sample, Treatment, analysis_group)
      ),
    by = c("sample", "Treatment", "analysis_group")
  ) %>%
  
  # Calculate extraction-induced changes
  mutate(
    COO_change      = COO_total_after - COO_total_before,
    carbonyl_change = carbonyl_after - carbonyl_before,
    aromatic_change = aromatic_after - aromatic_before,
    CH_change       = Aliphatic_total_after - Aliphatic_total_before
  )


# inspect
band_compare_long <- band_compare %>%
  # Filter out NA treatments if needed
  filter(!is.na(Treatment), Treatment != "NA") %>%
  pivot_longer(
    cols = ends_with("_change"),
    names_to = "functional_group",
    values_to = "change"
  ) %>%
  # Clean up column names (removes "_change" for cleaner facet titles)
  mutate(functional_group = gsub("_change$", "", functional_group))

band_compare_long <- band_compare %>%
  filter(!is.na(analysis_group)) %>%
  pivot_longer(
    cols = ends_with("_change"),
    names_to = "functional_group",
    values_to = "change"
  ) %>%
  mutate(functional_group = gsub("_change$", "", functional_group))


band_compare_filtered <- band_compare_long %>%
  filter(!is.na(analysis_group))

# Generate compact letter display per functional group
get_cld <- function(df) {
  dt <- FSA::dunnTest(change ~ factor(analysis_group), data = df, method = "bh")$res
  p_vals <- dt$P.adj
  names(p_vals) <- gsub(" ", "", dt$Comparison)
  
  cld <- multcompLetters(p_vals)$Letters
  data.frame(
    analysis_group = names(cld),
    letter = as.character(cld),
    stringsAsFactors = FALSE
  )
}

cld_df <- band_compare_filtered %>%
  group_by(functional_group) %>%
  group_modify(~ get_cld(.x)) %>%
  ungroup()

# Position letters slightly above mean + SE
pos_df <- band_compare_filtered %>%
  group_by(functional_group, analysis_group) %>%
  summarise(
    y_pos = mean(change, na.rm = TRUE) + (sd(change, na.rm = TRUE) / sqrt(n())) + 0.00005,
    .groups = "drop"
  )

cld_df <- left_join(cld_df, pos_df, by = c("functional_group", "analysis_group"))

# Plot with statistical letters
ggplot(band_compare_filtered, aes(x = analysis_group, y = change)) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "black") +
  geom_text(
    data = cld_df,
    aes(x = analysis_group, y = y_pos, label = letter),
    vjust = -0.5,
    size = 3.5,
    fontface = "bold"
  ) +
  facet_wrap(~ functional_group, scales = "free_y") +
  theme_bw() +
  labs(
    x = "Rock amendment",
    y = "Change in band area after extraction",
    title = "SOM Functional Group Changes by Treatment (Dunn test, p < 0.05)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  )



# ---------------------------------------------------------
# HÜHNERBERG DOSE-RESPONSE ANALYSIS
# ---------------------------------------------------------

huhnerberg_summary <- band_compare %>%
  filter(
    str_detect(Treatment, "^Huhnerberg_")
  ) %>%
  group_by(Dose) %>%
  summarise(
    n = n(),
    
    across(
      ends_with("_change"),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        se   = ~ sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x)))
      ),
      .names = "{.col}_{.fn}"
    ),
    
    .groups = "drop"
  )

huhnerberg_long <- band_compare %>%
  filter(
    str_detect(Treatment, "^Huhnerberg_")
  ) %>%
  pivot_longer(
    cols = ends_with("_change"),
    names_to = "functional_group",
    values_to = "change"
  ) %>%
  mutate(
    functional_group = gsub("_change$", "", functional_group)
  )

ggplot(
  huhnerberg_long,
  aes(x = Dose, y = change)
) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 3
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 1.5
  ) +
  facet_wrap(
    ~ functional_group,
    scales = "free_y"
  ) +
  theme_bw() +
  labs(
    x = expression("Hühnerberg application rate (t ha"^-1*")"),
    y = "Change in band area after extraction",
    title = "Spectral changes across the Hühnerberg dose-response gradient"
  )

```

## Interpration

**Control and lime behave same**
Both control and lime treatments undergo loss of band areas across almost all functional groups after extraction (COO, carbonyl and CH). Liming doesn't change extractability of SOM. Cation extraction results in loss of organic compounds that were loosely bound via weak cation bridging (Ca, Mg).

**Bolsdorfer and Hühnerberg rock dusts mitigate SOM loss**
Almost no aliphatic loss and reduced COO loss. Silicate rock dusts introduce reactive mineral surfaces (silicates, sesquioxides) that form stronger complexes with SOM that resist being stripped during cation extraction. Eifelgold only had intermediate effect.

**Dose-response experiment**
Increase in SOM extraction loss up to 30 t/ha, perhaps the increasing amount of cation bridges resulted in a bigger potential pool of NH4OAc-extractable SOM. Then at 50 t/ha, a critical saturation threshold may have been reached that resulted in formation of secondary minerals that formed organo-mineral complexes that are resistant to extraction  (or phsyical entrapment occurred). 