---
title: "ATR-FTIR Preliminary Analysis Update"
author: "Teneille Nel"
date: "`r Sys.Date()`"
format: html
editor: visual
---

## Overview: ATR Preliminary Findings

ATR-FTIR measurements of soil samples were taken before and after extractions of base cations (1M NH4AOc, pH = 7) to evaluate changes in organo-mineral environments between the control and rock powder treatments.

ATR measurements were imported to R and the data was cleaned by baseline correction by asymmetric least squares algorithm (improves signal dampened by noise and spectral drift), normalization (relative peak intensities to account for different sample thickness), Savitzky-Golay derivatization (to detect overlapping peaks due to absorption by different chemical compounds) and averaging of replicate scans. Difference curves were calculated to identify the changes in spectral signatures before and after extraction. Peak areas were also calculated under peaks across spectral ranges associated with functional groups typically found in carbohydrates, carbonates, silicates and organic bonds and the % change in these areas before and after extraction was calculated.

------------------------------------------------------------------------

## Preliminary Spectral Visualizations

The code generates plots comparing the pre- and post-extraction spectral changes for the Control (sample 1 and 18) vs. Bolsdorfer rock powder treatments (sample 4 and 20).

```{r}
#| echo: false
#| message: false
#| label: ATR comparison results
#| warning: FALSE

library(stringr)
library(reshape)
library(data.table)
library(nc)
library(baseline)
library(ggplot2)
library(tidyr)
library(pracma)
library(dplyr)
library(prospectr)


# ---------------------------------------------------------
# PROCESS SPECTRA
# ---------------------------------------------------------

process_spectra <- function(path){
  
  # ---- import ----
  long_table <- data.table(
    filename = Sys.glob(path)
  )[by = filename,
    j = fread(filename, col.names = c("wavenumber", "absorbance"))
  ]
  
  # ---- metadata ----
  meta <- nc::capture_first_vec(
    long_table$filename,
    sample_no = "\\d{1,2}",
    rep_no = "[a|b|c]{1,2}",
    nomatch.error = FALSE
  )
  
  meta_long <- cbind(long_table, meta)
  
  meta_long <- meta_long %>%
    mutate(
      wavenumber = round(wavenumber, 0),
      sample_rep = paste(sample_no, rep_no, sep = "_")
    ) %>%
    rename(
      wave = wavenumber,
      abs = absorbance,
      sample = filename
    )
  
  # ---------------------------------------------------------
  # BUILD WIDE MATRIX (raw)
  # ---------------------------------------------------------
  spec_wide <- dcast(meta_long, sample_rep ~ wave, value.var = "abs")
  
  ids <- spec_wide$sample_rep
  spec_mat <- as.matrix(spec_wide[, -1, with = FALSE])
  waves <- as.numeric(colnames(spec_wide)[-1])
  
  # ---------------------------------------------------------
  # BASELINE CORRECTION
  # ---------------------------------------------------------
  bl <- baseline::baseline(spec_mat, method = "als")
  spec_corr <- getCorrected(bl)
  
  corr_mat <- as.data.frame(spec_corr)
  corr_mat$sample_rep <- ids
  
  corr_long <- data.table::melt(
    setDT(corr_mat),
    id.vars = "sample_rep",
    variable.name = "wave",
    value.name = "abs_corr"
  )
  
  corr_long[, wave := as.numeric(as.character(wave))]
  
  # ---------------------------------------------------------
  # DEFINE COMMON GRID HERE
  # ---------------------------------------------------------
  common_grid <- sort(unique(corr_long$wave))
  
  # force consistent grid ordering
  corr_long <- corr_long %>%
    filter(wave %in% common_grid)
  
  # ---------------------------------------------------------
  # PEAK DETECTION (raw derivative space)
  # ---------------------------------------------------------
  
  spec_deriv <- prospectr::savitzkyGolay(
    X = spec_mat,
    p = 2,
    w = 11,
    m = 1
  )
  
  detect_peaks <- function(x, y, threshold = 0.0005){
    
    dy <- diff(y)
    peak_idx <- which(diff(sign(dy)) == -2) + 1
    peak_idx <- peak_idx[abs(y[peak_idx]) > threshold]
    
    data.frame(
      wave = x[peak_idx],
      intensity = y[peak_idx],
      index = peak_idx
    )
  }
  
  x <- waves
  
  peak_list <- lapply(seq_len(nrow(spec_deriv)), function(i){
    
    res <- detect_peaks(x, spec_deriv[i, ])
    
    if(is.null(res) || nrow(res) == 0) return(NULL)
    
    data.frame(
      sample_rep = ids[i],
      res
    )
  })
  
  peaks <- bind_rows(peak_list)
  
  list(
    raw = meta_long,
    corrected = corr_long,
    peaks = peaks,
    grid = common_grid
  )
}

# ---------------------------------------------------------
# RUN
# ---------------------------------------------------------

before <- process_spectra("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/rocmics/dpt_files/prelim_before/*.dpt")
after  <- process_spectra("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/rocmics/dpt_files/prelim_after/*.dpt")

# ---------------------------------------------------------
# FILTER REGION
# ---------------------------------------------------------

before_filt <- before$corrected %>%
  filter(wave >= 800, wave <= 3000)

after_filt <- after$corrected %>%
  filter(wave >= 800, wave <= 3000)

# ---------------------------------------------------------
# NORMALIZATION
# ---------------------------------------------------------

normalize_spectra <- function(df){
  
  df %>%
    group_by(sample_rep) %>%
    mutate(
      abs_norm = abs_corr / sqrt(sum(abs_corr^2, na.rm = TRUE))
    ) %>%
    ungroup()
}

before_filt <- normalize_spectra(before_filt)
after_filt  <- normalize_spectra(after_filt)

# ---------------------------------------------------------
# AVERAGING 
# ---------------------------------------------------------

before_avg <- before_filt %>%
  mutate(sample = sub("_[a-z]+$", "", sample_rep)) %>%
  group_by(sample, wave) %>%
  summarise(abs_corr = mean(abs_norm), .groups = "drop")

after_avg <- after_filt %>%
  mutate(sample = sub("_[a-z]+$", "", sample_rep)) %>%
  group_by(sample, wave) %>%
  summarise(abs_corr = mean(abs_norm), .groups = "drop")

# ---------------------------------------------------------
# SECOND DERIVATIVE
# ---------------------------------------------------------

compute_second_derivative <- function(df){
  
  samples <- unique(df$sample)
  grid <- sort(unique(df$wave))
  n <- length(grid)
  
  out <- list()
  
  for(s in samples){
    
    tmp <- df %>%
      filter(sample == s) %>%
      arrange(wave)
    
    y <- tmp$abs_corr
    
    y2 <- savitzkyGolay(
      matrix(y, nrow = 1),
      p = 3,
      w = 11,
      m = 2
    )[1, ]
    
    # -------------------------------------------------------
    # ALIGNMENT FIX: PAD TO FULL LENGTH
    # -------------------------------------------------------
    
    if(length(y2) != n){
      
      pad <- rep(NA, n)
      
      offset <- floor((n - length(y2)) / 2)
      
      pad[(offset + 1):(offset + length(y2))] <- y2
      
      y2 <- pad
    }
    
    out[[s]] <- data.frame(
      sample = s,
      wave = grid,
      deriv2 = as.numeric(y2)
    )
  }
  
  bind_rows(out)
}
before_deriv2 <- compute_second_derivative(before_avg)
after_deriv2  <- compute_second_derivative(after_avg)

# plot overlaid sample spectra

samples <- unique(before_avg$sample)

for(s in samples){
  
  overlay <- bind_rows(
    before_avg %>%
      filter(sample == s) %>%
      mutate(dataset = "Before"),
    
    after_avg %>%
      filter(sample == s) %>%
      mutate(dataset = "After")
  )
  
  p <- ggplot(overlay,
              aes(wave,
                  abs_corr,
                  colour = dataset)) +
    geom_line(size = 1) +
    scale_x_reverse() +
    theme_bw() +
    coord_cartesian(xlim = c(3000, 800)) +
    labs(
      title = paste("Sample", s),
      x = "Wavenumber (cm-1)",
      y = "Corrected absorbance"
    )
  
  print(p)
  
}

# --------------------------------------------------
# Difference spectra
# after - before
# --------------------------------------------------

diff_spec <- before_avg %>%
  rename(before = abs_corr) %>%
  left_join(
    after_avg %>%
      rename(after = abs_corr),
    by = c("sample", "wave")
  ) %>%
  mutate(diff = after - before)

# --------------------------------------------------
# Plot all samples
# --------------------------------------------------

ggplot(diff_spec,
       aes(wave, diff, colour = factor(sample))) +
  
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  
  geom_line(size = 0.8) +
  
  scale_x_reverse() +
  
  theme_bw() +
  
  labs(
    title = "Difference spectra (After − Before)",
    x = expression(Wavenumber~(cm^{-1})),
    y = "Difference absorbance",
    colour = "Sample"
  ) +
  
  coord_cartesian(
    xlim = c(1800, 800)
  )

ggplot(diff_spec,
       aes(wave, diff)) +
  
  geom_hline(yintercept = 0,
             linetype = "dashed") +
  
  geom_line(size = 0.8) +
  
  scale_x_reverse() +
  
  facet_wrap(~sample,
             ncol = 2,
             scales = "free_y") +
  
  theme_bw() +
  
  labs(
    title = "Difference spectra (After − Before)",
    x = expression(Wavenumber~(cm^{-1})),
    y = "Difference absorbance"
  ) +
  
  coord_cartesian(
    xlim = c(1800, 800)
  )
```

## Band region integrations

The code generates changes in areas under absorbance spectra associated with functional groups involved in interactions with cations.

```{r}
#| echo: false
#| message: false
#| label: ATR integration results
#| warning: FALSE

# ---------------------------------------------------------
# Function to integrate spectral regions
# ---------------------------------------------------------

integrate_bands <- function(df){
  
  bands <- list(
    COO_sym  = c(1380,1425),   # COO- symmetric stretch
    COO_asym = c(1580,1650),   # COO- asymmetric stretch
    Silicate = c(1000,1120),   # silicates
    Carbonate_bend = c(870,880),     # carbonate bend
    carbonate_stretch = c(1415,1480),
    double_bond = c(1510,1580), #aromatic & amide
    polysaccharide = c(1030-1080) # C-O and mineral-bound carbohydrates
  )
  
  results <- lapply(unique(df$sample), function(s){
    
    sub <- df %>% filter(sample == s)
    
    out <- data.frame(sample = s)
    
    for(b in names(bands)){
      
      rng <- bands[[b]]
      
      tmp <- sub %>%
        filter(wave >= rng[1],
               wave <= rng[2]) %>%
        arrange(wave)
      
      area <- trapz(tmp$wave, tmp$abs_corr)
      
      out[[b]] <- area
    }
    
    out
  })
  
  bind_rows(results)
}

# ---------------------------------------------------------
# Calculate integrated areas
# ---------------------------------------------------------

before_areas <- integrate_bands(before_avg) %>%
  mutate(dataset = "before")

after_areas <- integrate_bands(after_avg) %>%
  mutate(dataset = "after")

# ---------------------------------------------------------
# Combine and calculate extraction effect
# ---------------------------------------------------------

band_compare <- before_areas %>%
  select(-dataset) %>%
  rename_with(~paste0(.x,"_before"), -sample) %>%
  left_join(
    after_areas %>%
      select(-dataset) %>%
      rename_with(~paste0(.x,"_after"), -sample),
    by = "sample"
  ) %>%
  mutate(
    COO_sym_change =
      100*(COO_sym_after - COO_sym_before)/COO_sym_before,
    
    COO_asym_change =
      100*(COO_asym_after - COO_asym_before)/COO_asym_before,
    
    Silicate_change =
      100*(Silicate_after - Silicate_before)/Silicate_before,
    
    Carbonate_bend_change =
      100*(Carbonate_bend_after - Carbonate_bend_before)/Carbonate_bend_before,
    
    carbonate_stretch_change =
      100*(carbonate_stretch_after - carbonate_stretch_before)/carbonate_stretch_before,
    
    double_bond_change =
      100*(double_bond_after - double_bond_before)/double_bond_before,
    
    polysaccharide_change =
      100*(polysaccharide_after - polysaccharide_before)/polysaccharide_before  
    )

# inspect
band_compare

```

------------------------------------------------------------------------

## Key Results

- **Control Treatment Dynamics:**
  - Demonstrated a higher abundance of **labile cation-organic associations**.
  - Extraction induced a highly consistent shift in the carboxylate environment, characterized by a **decrease in the** $COO^-$ asymmetric stretching region and a corresponding **increase in the** $COO^-$ symmetric stretching region.
- **Rock Powder Treatment Stability:**
  - Significantly **reduced the sensitivity of the carboxylate environment** to chemical extraction.
  - Indicates the formation of **more stable, less extractable organo-mineral complexes**.
- **Mineral Matrix Preservation:**
  - The **silicate region remained largely stable** across extractions, confirming that the underlying mineral matrix was conserved.

## Interpretations & Next Steps

### Primary Interpretation: Enhanced Organo-Mineral Passivation

If the rock powder successfully weathered and released structural cations ($\text{Ca}^{2+}$, $\text{Mg}^{2+}$, $\text{Fe}^{3+}$, or $\text{Al}^{3+}$), organic matter likely became tightly associated with newly formed mineral surfaces or precipitated mineral phases. Because these bonds are no longer primarily held together by easily exchangeable ions, removing the exchangeable cation pool did not drastically disrupt the carboxylate spectral environment.

### Alternative Hypotheses & Experimental Verification

1.  **Cation Saturation/Incomplete Extraction**
    - *Hypothesis:* Weathering may have released a pool of structural $\text{Ca}^{2+}$ or $\text{Mg}^{2+}$ so large that the current extraction protocol could not completely deplete it, resulting in a deceptively small net spectral change.
    - *Validation:* We can rule this out by running a **longer extraction time** for the full sample set (the preliminary test only used a 30-minute extraction window).
2.  **Physical Aggregation & Mass Transfer Limitations**
    - *Hypothesis:* The rock powder treatment may promote physical aggregation, physically shielding reactive surfaces from interacting with the extraction solution.
    - *Validation:* We will explicitly test this physical stabilization mechanism via the upcoming **sonication experiment**.

------------------------------------------------------------------------
