---
title: "dcb_oxalate_combined"
format: html
editor: visual
---

## Overview

Teneille's full project repository is available at <https://github.com/Teneille97/rocmics>.

We analysed crystalline and non-crystalline metal(loid)s (Fe, Al, Mn and Si) in an enhanced rock weathering field trial in Malle.

The DCB extraction is designed to reduce crystalline sesquioxides and the oxalate extraction is highly acidic and can dissolve short-order range (amorphous, non-crystalline) aluminosilicates. 

## DCB- and oxalate-extractable metal(loid) content of different treatments

```{r}
#| echo: false
#| message: false
#| label: metal(loid)s comparison
#| warning: FALSE
#| 
# ==============================================================================
# 1. LIBRARIES & SETUP
# ==============================================================================
library(tidyverse)
library(here)
library(emmeans)
library(multcomp)
library(multcompView)
library(ggplot2)
library(dplyr)
library(broom)

# Set global ggplot theme
theme_set(theme_bw())

# ==============================================================================
# 2. METADATA PROCESSING
# ==============================================================================
metadata <- read.csv(here("csv_files", "treatment_names.csv")) %>%
  mutate(
    analysis_group = case_when(
      Treatment == "Control" ~ "Control",
      Treatment == "Eifelgold_50" ~ "Eifelgold_50",
      Treatment == "Bolsdorfer_50" ~ "Bolsdorfer_50",
      Treatment == "Lime_2" ~ "Lime_2",
      Treatment == "Huhnerberg_50" ~ "Huhnerberg_50",
      TRUE ~ NA_character_
    )
  )

# ==============================================================================
# 3. IMPORT & CLEAN OXALATE DATA (mg g^-1)
# ==============================================================================
oxalate_raw <- read.csv(here("csv_files", "oxalate_sesq_feb2026.csv"))

oxalate_long <- cbind(oxalate_raw, metadata) %>%
  separate(Treatment, into = c("Tmt", "App_rate"), sep = "_", fill = "right") %>%
  mutate(
    App_rate = replace_na(App_rate, "0"),
    Tmt = factor(Tmt),
    App_rate = factor(App_rate, levels = c("0", "2", "4", "8", "12", "20", "30", "50")),
    Method = "Oxalate"
  ) %>%
  dplyr::select(sample, analysis_group, Tmt, App_rate, Method,
                Fe = Fe..mg.g.1.soil.,
                Al = Al..mg.g.1.soil.,
                Mn = Mn..mg.g.1.soil.,
                Si = Si..mg.g.1.soil.) %>%
  pivot_longer(
    cols = c(Fe, Al, Mn, Si),
    names_to = "Element",
    values_to = "Concentration"
  )

# ==============================================================================
# 4. IMPORT & CLEAN DCB DATA (Converted to mg g^-1: % * 10)
# ==============================================================================
dcb_raw <- read.csv(here("csv_files", "DCB_Feb2026.csv"))

# Note: Converting elemental % to mg g^-1 by multiplying by 10 (1% = 10 mg/g)
dcb_long <- cbind(dcb_raw, metadata) %>%
  separate(Treatment, into = c("Tmt", "App_rate"), sep = "_", fill = "right") %>%
  mutate(
    App_rate = replace_na(App_rate, "0"),
    Tmt = factor(Tmt),
    App_rate = factor(App_rate, levels = c("0", "2", "4", "8", "12", "20", "30", "50")),
    Method = "DCB",
    # Column conversions if elemental % columns exist
    Fe = if("X.Fe" %in% names(.)) X.Fe * 10 else NA_real_,
    Al = X.Al * 10,
    Mn = X.Mn * 10,
    Si = X.Si * 10
  ) %>%
  dplyr::select(sample, analysis_group, Tmt, App_rate, Method, Fe, Al, Mn, Si) %>%
  pivot_longer(
    cols = c(Fe, Al, Mn, Si),
    names_to = "Element",
    values_to = "Concentration"
  )

# ==============================================================================
# 5. UNIFIED DATASET & METRIC CALCULATIONS (DIFFERENCES & RATIOS)
# ==============================================================================
# Combined long dataset for extractions
combined_extractions <- bind_rows(oxalate_long, dcb_long)

# Reshape wide to compute sample-matched ratios and differences
extraction_metrics <- combined_extractions %>%
  pivot_wider(
    names_from = Method,
    values_from = Concentration
  ) %>%
  mutate(
    Diff_DCB_Ox = DCB - Oxalate,           # Crystalline/Residual fraction (e.g., Fe_d - Fe_o)
    Ratio_Ox_DCB = Oxalate / DCB           # Amorphous/Active fraction ratio (e.g., Fe_o / Fe_d)
  )

# Convert metrics to long format for easy visualization
metrics_long <- extraction_metrics %>%
  pivot_longer(
    cols = c(Diff_DCB_Ox, Ratio_Ox_DCB),
    names_to = "Metric",
    values_to = "Value"
  )

# ==============================================================================
# 6. COMBINED PLOTS: OXALATE VS DCB (SHARED SCALE)
# ==============================================================================

# Custom Color Palette
method_colors <- c("Oxalate" = "#E69F00", "DCB" = "#56B4E9")

# --- 50 t/ha Treatments Comparison (ANOVA & Tukey Letters) ---
plot50_combined <- combined_extractions %>%
  filter(analysis_group %in% c("Control", "Bolsdorfer_50", "Eifelgold_50", "Huhnerberg_50", "Lime_2")) %>%
  mutate(analysis_group = factor(analysis_group, levels = c("Control", "Bolsdorfer_50", "Eifelgold_50", "Huhnerberg_50", "Lime_2")))

# Generate Tukey Letters per Element and Extraction Method
letters_df_combined <- plot50_combined %>%
  group_by(Element, Method) %>%
  group_modify(~{
    mod <- aov(Concentration ~ analysis_group, data = .x)
    emm <- emmeans(mod, ~ analysis_group)
    cld_res <- cld(emm, Letters = letters, adjust = "tukey")
    
    cld_res %>%
      as.data.frame() %>%
      dplyr::select(analysis_group, .group) %>%
      mutate(
        .group = str_trim(.group),
        y = max(.x$Concentration, na.rm = TRUE) * 1.08
      )
  }) %>%
  ungroup()

ggplot(plot50_combined, aes(x = analysis_group, y = Concentration, fill = Method)) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.7,
    position = position_dodge(0.8)
  ) +
  geom_point(
    aes(color = Method),
    position = position_jitterdodge(
      jitter.width = 0.1,
      dodge.width = 0.8
    ),
    size = 1.5
  ) +
  geom_text(
    data = letters_df_combined,
    aes(
      x = analysis_group,
      y = y,
      label = .group,
      group = Method
    ),
    position = position_dodge(0.8),
    size = 3.5,
    vjust = 0
  ) +
  facet_grid(Element ~ Method, scales = "free_y") +
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  scale_fill_manual(values = method_colors) +
  scale_color_manual(values = method_colors) +
  labs(
    x = "",
    y = expression(paste("Concentration (mg g"^-1, " soil)")),
    title = "50 t/ha Group Comparisons with Post-Hoc Tukey Groups"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

# --- Dose-Response (Huhnerberg vs Control) ---
dose_resp_combined <- combined_extractions %>%
  filter(Tmt == "Control" | Tmt == "Huhnerberg") %>%
  mutate(
    Treatment = ifelse(Tmt == "Control", "Control", paste0(App_rate, " t ha^-1")),
    Treatment = factor(Treatment, levels = c("Control", "2 t ha^-1", "4 t ha^-1", "8 t ha^-1", "12 t ha^-1", "20 t ha^-1", "30 t ha^-1", "50 t ha^-1"))
  )

ggplot(dose_resp_combined, aes(x = Treatment, y = Concentration, fill = Method)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(0.8)) +
  geom_point(aes(color = Method), position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8), size = 1.8) +
  facet_grid(Element ~ ., scales = "free_y") +
  scale_fill_manual(values = method_colors) +
  scale_color_manual(values = method_colors) +
  labs(
    x = "",
    y = expression(paste("Concentration (mg g"^-1, " soil)")),
    title = "Dose-Response Comparison (Huhnerberg)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

## Ratio comparison

We then compared ratios of amorphous to crystalline elements among different treatments. 

```{r}
#| echo: false
#| message: false
#| label: ratio comparison
#| warning: FALSE
#| 
metrics_50 <- extraction_metrics %>%
  filter(analysis_group %in% c("Control", "Bolsdorfer_50", "Eifelgold_50", "Huhnerberg_50", "Lime_2")) %>%
  mutate(analysis_group = factor(analysis_group, levels = c("Control", "Bolsdorfer_50", "Eifelgold_50", "Huhnerberg_50", "Lime_2")))


# 1. Filter metrics data for the 50 t/ha comparison group
ratio_50 <- extraction_metrics %>%
  filter(analysis_group %in% c("Control", "Bolsdorfer_50", "Eifelgold_50", "Huhnerberg_50", "Lime_2")) %>%
  mutate(analysis_group = factor(analysis_group, levels = c("Control", "Bolsdorfer_50", "Eifelgold_50", "Huhnerberg_50", "Lime_2")))

# 2. Calculate Tukey HSD compact letter displays (CLD) per Element for Ratio_Ox_DCB
letters_ratio_df <- ratio_50 %>%
  group_by(Element) %>%
  group_modify(~{
    # Run ANOVA model on Ratio_Ox_DCB
    mod <- aov(Ratio_Ox_DCB ~ analysis_group, data = .x)
    
    # Estimate marginal means
    emm <- emmeans(mod, ~ analysis_group)
    
    # Compute Tukey pairwise comparisons and letter groups
    cld_results <- cld(
      emm,
      Letters = letters,
      adjust = "tukey"
    )
    
    # Format letters and calculate label y-position slightly above max ratio value
    cld_results %>%
      as.data.frame() %>%
      dplyr::select(analysis_group, .group) %>%
      mutate(
        .group = str_trim(.group),
        y = max(.x$Ratio_Ox_DCB, na.rm = TRUE) * 1.10
      )
  }) %>%
  ungroup()

# 3. Plot Amorphous Fraction Ratio with Tukey Letters
ggplot(ratio_50, aes(x = analysis_group, y = Ratio_Ox_DCB)) +
  geom_boxplot(
    outlier.shape = NA,
    fill = "grey92",
    color = "grey30"
  ) +
  geom_jitter(
    width = 0.15,
    size = 2,
    alpha = 0.8,
    color = "#2B5C8F"
  ) +
  geom_text(
    data = letters_ratio_df,
    aes(
      x = analysis_group,
      y = y,
      label = .group
    ),
    inherit.aes = FALSE,
    size = 4.5,
    fontface = "bold"
  ) +
  facet_grid(Element ~ ., scales = "free_y") +
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  labs(
    x = "",
    y = expression(paste("Active Ratio (Oxalate / DCB)")),
    title = "Amorphous Fraction Ratio (Oxalate / DCB) with Tukey HSD Groups"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(face = "bold", size = 13)
  )
# ==============================================================================
# Dose-response ratio plot (Hühnerberg)
# No statistical comparisons because there are no replicates
# ==============================================================================

dose_resp_ratio <- extraction_metrics %>%
  filter(Tmt == "Control" | Tmt == "Huhnerberg") %>%
  mutate(
    Treatment = ifelse(
      Tmt == "Control",
      "Control",
      paste0(App_rate, " t ha^-1")
    ),
    Treatment = factor(
      Treatment,
      levels = c(
        "Control",
        "2 t ha^-1",
        "4 t ha^-1",
        "8 t ha^-1",
        "12 t ha^-1",
        "20 t ha^-1",
        "30 t ha^-1",
        "50 t ha^-1"
      )
    )
  )


ggplot(
  dose_resp_ratio,
  aes(x = Treatment, y = Ratio_Ox_DCB, group = 1)
) +
  geom_point(
    size = 3,
    alpha = 0.9
  ) +
  facet_grid(
    Element ~ .,
    scales = "free_y"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  labs(
    x = "",
    y = expression(paste("Active Ratio (Oxalate / DCB)")),
    title = "Dose-Response Comparison (Hühnerberg)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 10
    ),
    strip.text = element_text(
      size = 11,
      face = "bold"
    ),
    plot.title = element_text(
      face = "bold",
      size = 13
    )
  )
```
