---
title: "Density fractionation initial results"
format: html
author: "Teneille Nel"
date: "`r Sys.Date()`"
editor: visual
---

## Density fractionation procedure

Teneille's data analysis repository is available at <https://github.com/Teneille97/rocmics>.
I performed density fractionation by density fractionation using SPT and ca. 10 g soil at Max Planck Institute for Biogeochemistry.

## SOM pools as proportions of bulk

Mass-based pools recovered from density fractionation were as follows:

```{r}
#| echo: false
#| message: false
#| label: Proportions of SOM pools
#| warning: FALSE

library(tidyverse)
library(here)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyr)
library(stringr)


# import
density_fractionation <- read.csv(here("csv_files", "density_fractionation_data_final.csv")) #replace with final csv version after replacing originals with redo's
moisture_content <- read.csv(here("csv_files", "moisture_content_density_fractionation.csv"))
rbf_masses <- read.csv(here("csv_files", "rbf_masses_density_fractionation.csv"))
treatment_names <- read.csv(here("csv_files", "treatment_names.csv"))
CN_data<-read.csv(here("csv_files", "total_CN_MAOM_bulk.csv"))

# clean + join + compute
density_fractionation <- density_fractionation %>%
  
  # --- clean lookup tables ---
  left_join(
    moisture_content %>%
      rename(moisture_content = Water.content..m.m.1.),
    by = c("Plot" = "Sample",
           "Sampling_year" = "Sample.year")
  ) %>%
  
  left_join(
    rbf_masses %>%
      rename(rbf = RBF., mass_rbf = mass..g.),
    by = c("rbf.." = "rbf")
  ) %>%
  
  left_join(
    treatment_names %>%
      rename(Plot = sample),
    by = "Plot"
  ) %>%
  
  # --- drop junk columns safely ---
  select(-any_of(c("mass.rbf", "moisture.content", "recovery....",
                   "som.fraction....", "X", "X.1", "X.2", "Sample.date"))) %>%
  
  # --- fix treatment (2023 = Control) ---
  mutate(
    Treatment = if_else(Sampling_year == 2023, "Control", Treatment)
  ) %>%
  
  # --- split treatment ---
  separate(Treatment, into = c("Tmt", "App_rate"), sep = "_", fill = "right") %>%
  mutate(
    App_rate = replace_na(App_rate, "0"),
    Tmt = factor(Tmt),
    App_rate = factor(App_rate)
  ) %>%
  
  # --- numeric fixes + calculations ---
  mutate(
    mass.som.on.filter.corr = as.numeric(mass.som.on.filter.corr),
    mass.som.on.filter.corr = replace_na(mass.som.on.filter.corr, 0),
    
    mass.soil.corr = mass.soil * (1 - moisture_content),
    
    mass.som.fraction =
      mass.rbf...som.fraction.freeze.dried +
      mass.som.on.filter.corr -
      mass_rbf,
    
    SOM_fraction = factor(SOM_fraction)
  )

# calculate total mass som fractions per sample
density_fractionation <- density_fractionation %>%
  # 1. Sort by Sampling_year, then Plot
  arrange(Sampling_year, Plot) %>%
  
  # 2. Group by the unique identifier for a physical soil sample
  group_by(Sampling_year, Plot) %>%
  
  # 3. Calculate the sum of mass.som.fraction and assign it to the column
  mutate(total.mass.som.fractions = sum(mass.som.fraction, na.rm = TRUE)) %>%
  
  # 4. Ungroup 
  ungroup()


#calculate recovery
density_fractionation$recovery_percent<-100*density_fractionation$total.mass.som.fractions/density_fractionation$mass.soil.corr

# Calculate the proportion of each fraction within its own sample
df_proportions <- density_fractionation %>%
  mutate(
    fraction_prop = (mass.som.fraction / total.mass.som.fractions) * 100
  )
# add CN data - bulk 2026 & MAOM

# Convert CN data from long to wide format
CN_wide <- CN_data %>%
  select(Sample.name, Parameter, Result....) %>%
  pivot_wider(
    names_from = Parameter,
    values_from = Result....
  )

# create bulk lookup table
CN_bulk <- CN_wide %>%
  filter(str_detect(Sample.name, "Bulk")) %>%
  mutate(
    core_id = str_replace(Sample.name, "_Bulk$", "")
  ) %>%
  select(core_id, Ctotal, Ntotal) %>%
  rename(C_bulk = Ctotal,
         N_bulk = Ntotal)

# create matching id in df proportions
df_proportions <- df_proportions %>%
  mutate(
    core_id = str_replace(sample.ID, "_(fPOM|oPOM|MAOM)$", "")
  )

# Join to df_proportions by sample ID
df_proportions <- df_proportions %>%
  left_join(CN_wide, by = c("sample.ID" = "Sample.name")) %>%
  left_join(CN_bulk, by = "core_id")

df_proportions$CN<-df_proportions$Ctotal/df_proportions$Ntotal
df_proportions$CN_bulk<-df_proportions$C_bulk/df_proportions$N_bulk

# add CN data - bulk 2025

CN_data_nov2025 <- read.csv(here("csv_files", "CN_nov2025.csv"), header = TRUE)
CN_data_nov2025 <- CN_data_nov2025[1:63,]

clean_data <- function(df) {
  df %>%
    mutate(across(c(Treatment, Dose_response_exp.), as.factor)) %>%
    separate(Treatment, into = c("Tmt", "App_rate"), sep = "_") %>%
    mutate(
      App_rate = replace_na(App_rate, "0"),
      Tmt = as.factor(Tmt)
    )%>%
    rename(
      N = X.N,
      C = X.C
    )
}

CN_data_nov2025 <- clean_data(CN_data_nov2025)
CN_data_nov2025$CN<-CN_data_nov2025$C/CN_data_nov2025$N

# Summarise across reps
CN_data_nov2025 <- CN_data_nov2025 %>%
  group_by(sample) %>%
  summarise(across(
    c(N, C, CN),
    list(mean = ~ mean(.x, na.rm = TRUE),
         se   = ~ sd(.x, na.rm = TRUE) / sqrt(n())),
    .names = "{.fn}_{.col}"),
    .groups = "drop")

CN_data_nov2025<-as.data.frame(CN_data_nov2025)
CN_data_nov2025$sample <- paste0(CN_data_nov2025$sample, "_2025_Bulk")

# Prepare 2025 bulk CN data
CN_2025 <- CN_data_nov2025 %>%
  mutate(core_id = str_remove(sample, "_Bulk$")) %>%
  select(
    core_id,
    mean_C,
    mean_N,
    mean_CN
  )

# Join and populate existing bulk columns
df_proportions <- df_proportions %>%
  left_join(CN_2025, by = "core_id") %>%
  mutate(
    C_bulk  = if_else(Sampling_year == 2025, mean_C, C_bulk),
    N_bulk  = if_else(Sampling_year == 2025, mean_N, N_bulk),
    CN_bulk = if_else(Sampling_year == 2025, mean_CN, CN_bulk)
  ) %>%
  select(-mean_C, -mean_N, -mean_CN)


# multiply %C with total C in sample for proportions (% of total C distributed in each pool)
df_proportions$Cprop<-(df_proportions$Ctotal*df_proportions$fraction_prop)/(df_proportions$C_bulk)
df_proportions$Nprop<-(df_proportions$Ntotal*df_proportions$fraction_prop)/(df_proportions$N_bulk)

#visualize SOM proportions (mass-based)
# 1. Prepare data (filter for replicates and ensure proportions are calculated)
plot_data_faceted <- df_proportions %>%
  mutate(fraction_prop = 100 * (mass.som.fraction / total.mass.som.fractions)) %>%
  group_by(Sampling_year, Tmt, App_rate) %>%
  filter(n_distinct(Plot) > 1) %>%
  ungroup()

# 2. Create the plot
ggplot(plot_data_faceted, aes(x = factor(Tmt), y = fraction_prop, fill = Tmt)) +
  # Using boxplots or points here is better than bars for comparing replicates
  geom_boxplot(alpha = 0.3, outlier.shape = NA) + 
  geom_jitter(width = 0.1, alpha = 0.7) +
  # Facet by Fraction (Rows) and Year (Columns)
  # 'free_y' is the magic ingredient here
  facet_grid(SOM_fraction ~ Sampling_year, scales = "free_y") +
  labs(
    title = "Comparison of SOM Fraction Proportions",
    subtitle = "Note: Y-axes differ by fraction to highlight POM variability",
    x = "Treatment",
    y = "Percentage of Total SOM (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none" # Hide legend since X-axis labels cover it
  )

```

The total C and N content of bulk soil and SOM fractions was measured by combustion. The %C of each pool was multiplied by the proportion of that pool as a fraction of the bulk to determine the distribution of bulk soil C among pools:

```{r}
#| echo: false
#| message: false
#| label: Distribution of C among pools
#| warning: FALSE



# plot C and N in SOM fraction as proportion of bulk 

#first C
ggplot(plot_data_faceted, aes(x = factor(Tmt), y = Cprop, fill = Tmt)) +
  # Using boxplots or points here is better than bars for comparing replicates
  geom_boxplot(alpha = 0.3, outlier.shape = NA) + 
  geom_jitter(width = 0.1, alpha = 0.7) +
  facet_grid(SOM_fraction ~ Sampling_year, scales = "free_y") +
  labs(
    title = "Comparison of SOM pool C as proportion of bulk C (%)",
    subtitle = "Note: Y-axes differ by fraction to highlight POM variability",
    x = "Treatment",
    y = "Proportion of Total SOM C"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none" # Hide legend since X-axis labels cover it
  )

#then N
ggplot(plot_data_faceted, aes(x = factor(Tmt), y = Nprop, fill = Tmt)) +
  # Using boxplots or points here is better than bars for comparing replicates
  geom_boxplot(alpha = 0.3, outlier.shape = NA) + 
  geom_jitter(width = 0.1, alpha = 0.7) +
  facet_grid(SOM_fraction ~ Sampling_year, scales = "free_y") +
  labs(
    title = "Comparison of SOM pool N as proportion of bulk N (%)",
    subtitle = "Note: Y-axes differ by fraction to highlight POM variability",
    x = "Treatment",
    y = "Proportion of Total SOM N"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none" # Hide legend since X-axis labels cover it
  )



```

The % change of C distribution was normalized to the control as (treatment-control)/control x 100%.

```{r}
#| echo: false
#| message: false
#| label: Normalized to control
#| warning: FALSE


# normalize by control in that year (z-score effect)
control_baseline <- df_proportions %>%
  mutate(fraction_prop = 100 * (mass.som.fraction / total.mass.som.fractions)) %>%
  filter(Tmt == "Control", App_rate == "0") %>%
  group_by(Sampling_year, SOM_fraction) %>%
  summarise(
    control_Cprop = mean(Cprop, na.rm = TRUE),
    control_Nprop = mean(Nprop, na.rm = TRUE),
    .groups = "drop"
  )

plot_data_pct <- df_proportions %>%
  mutate(fraction_prop = 100 * (mass.som.fraction / total.mass.som.fractions)) %>%
  
  filter(
    (Tmt == "Control"     & App_rate == "0") |
      (Tmt == "Lime"        & App_rate == "2") |
      (Tmt == "Bolsdorfer"  & App_rate == "50") |
      (Tmt == "Eifelgold"   & App_rate == "50") |
      (Tmt == "Huhnerberg"   & App_rate == "50")
  ) %>%
  
  left_join(control_baseline, by = c("Sampling_year", "SOM_fraction")) %>%
  
  mutate(
    Cprop_pct = ((Cprop - control_Cprop) / control_Cprop) * 100,
    Nprop_pct = ((Nprop - control_Nprop) / control_Nprop) * 100
  ) %>%
  
  group_by(Sampling_year, Tmt, App_rate) %>%
  filter(n_distinct(Plot) > 1) %>%
  ungroup()

ggplot(plot_data_pct, aes(x = factor(Tmt), y = Cprop_pct, fill = Tmt)) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  facet_grid(SOM_fraction ~ Sampling_year, scales = "free_y") +
  labs(
    title = "SOM pool C (% change vs control)",
    subtitle = "Control-normalised within year and SOM fraction",
    x = "Treatment",
    y = "% change in C proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(plot_data_pct, aes(x = factor(Tmt), y = Nprop_pct, fill = Tmt)) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  facet_grid(SOM_fraction ~ Sampling_year, scales = "free_y") +
  labs(
    title = "SOM pool N (% change vs control)",
    x = "Treatment",
    y = "% change in N proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
